---
title: "Additional Data"
output: html_notebook
---

```{r setup}

# working directory (should be /final-project-bald_eagle)
knitr::opts_knit$set(root.dir = normalizePath('../'))

# packages
library(tidyverse)
library(janitor)

```

This reads a saved csv of expenditures_ts, written from the code in our_data at commit eaa3885.

```{r expenditures_ts}

# loading data
raw_data <- read_csv('dataset/expenditures_ts.csv',
                     col_types = 'ccccdddddd')

```

For my preference, some cleaning up of names and columns.  Adding a binomial (Genus species) column to help with searching later; removing features that aren't helpful.

```{r cleaning}

# add binomial; remove unnecessary features
expen <- raw_data %>% 
  clean_names() %>% 
  mutate(binomial = paste(snakecase::to_any_case(genus, 'sentence'),
                          species,
                          sep = ' ')) %>% 
  select(class, binomial, fws_total:year)

# should be no NA
any(is.na(expen))

```

Date notes:
- There are some errors where species name doesn't make sense---mis-imported data?

A list of species with most data and most expenditures (both average and overall).  `number` is number of data points, `average_exp` is average expenditure, and `overall_exp` is total expenditure for the species.

```{r notable-species}

notable <- expen %>% 
  
  # combine subspecies per year, keep both average and total exp.
  group_by(class, binomial, year) %>% 
  summarize(average_subsp = mean(species_total),
            total_subsp = sum(species_total),
            .groups = 'keep') %>% 
  ungroup() %>% 
  
  # per-species insights
  group_by(class, binomial) %>% 
  summarize(number = n(),
            average_exp = mean(average_subsp),
            average_exp_tot = sum(average_subsp),
            overall_exp_avg = mean(total_subsp),
            overall_exp = sum(total_subsp),
            .groups = 'keep') %>% 
  mutate(class = ifelse(class == 'amphibias', 'amphibia', class))

```

Species with the most data:

```{r most-data}

# list
(most_data <- notable %>% 
  select(class:number) %>% 
  arrange(desc(number)))

# plot
ggplot(notable, aes(class, number, color = scale(average_exp_tot))) +
  geom_jitter(alpha = 0.7) +
  scale_color_gradient(low = 'lightblue', high = 'black') +
  labs(title = 'Classes with most data',
       x = 'Class',
       y = 'Number of years with data',
       color = 'Normalized\nexpenditure')

```

Species with the most average expenditure:

```{r most-avg}

notable %>% 
  select(class, binomial, average_exp) %>% 
  arrange(desc(average_exp)) %>% 
  head(10) %>% 
  .$binomial -> top_avgavg

notable %>% 
  select(class, binomial, overall_exp_avg) %>% 
  arrange(desc(overall_exp_avg)) %>% 
  head(10) %>% 
  .$binomial -> top_totavg

```

Species with the most total expenditure

```{r most-total}

notable %>% 
  select(class, binomial, overall_exp) %>% 
  arrange(desc(overall_exp)) %>% 
  head(10) %>% 
  .$binomial -> top_tottot

notable %>% 
  select(class, binomial, average_exp_tot) %>% 
  arrange(desc(average_exp_tot)) %>% 
  head(10) %>% 
  .$binomial -> top_avgtot

```

List of all species to pull additional data for.  Note that "Ca ca", "Snake stock", and "Snake found." are all non-species.

```{r list-species}

unique(c(top_avgavg, top_avgtot, top_totavg, top_tottot))

```

Common names of found species (and file names):
pallid sturgeon (Scaphirhynchus albus)
-pallid_sturgeon.csv
bull trout (Salvelinus confluentus)
-bull_trout.csv
rainbow trout (Oncorhynchus mykiss)
-rainbow_trout.csv

(salmon are mostly together)
Chinook salmon (Oncorhynchus tshawytsch)
sockeye salmon (Oncorhynchus nerka)
coho salmon (Oncorhynchus kisutch)
chum salmon (Oncorhynchus keta)
-salmon.pdf (different data than salmon.csv; needs scraping)
-salmon.csv
-chinook.pdf
-survival_chinook_steelhead.csv


Datasets not found/not looked for:
Steller sea lion
red-cockaded woodpecker
willow flycatcher
razorback sucker
Indiana bat
desert tortoise

For some preliminary analysis, a look into the pallid sturgeon.  Note that the survey data was from various surveys, so it was averaged per year (to avoid overweighting a year with more surveys).

```{r pallid-sturgeon}

raw_sturgeon <- read_csv('dataset/additional/pallid_sturgeon.csv',
                         col_types = 'ciccdddd')

# years surveyed per waterbody
raw_sturgeon %>% 
  clean_names() %>% 
  select(year, waterbody, count) %>%
  group_by(waterbody) %>% 
  summarize(num_years = length(unique(year)),
            .groups = 'keep')

# mean counts after outliers removed
sturgeon <- raw_sturgeon %>% 
  clean_names() %>% 
  filter(!waterbody %in% c('Fort Peck Dredge Cuts', 'Milk River')) %>% 
  select(year, count) %>% 
  group_by(year) %>% 
  summarize(avg_count = mean(count),
            .groups = 'keep') %>% 
  arrange(year)

# sturgeon expenditure
sturgeon_exp <- expen %>% 
  filter(binomial == 'Scaphirhynchus albus') %>%
  inner_join(sturgeon, by = 'year')

# plot expenditure vs sturgeon
ggplot(sturgeon_exp, aes(year)) +
  geom_smooth(aes(y = scale(avg_count)),
              color = 'red',
              se = F) +
  geom_smooth(aes(y = scale(species_total)),
              color = 'blue',
              se = F) +
  labs(title = 'Sturgeon expenditure and population',
       x = 'Year',
       y = 'Normalized expenditure/population counts')

# linear model
reg_sturgeon <- lm(avg_count ~ species_total,
                   data = sturgeon_exp)
summary(reg_sturgeon)
plot(reg_sturgeon)

# pretty badly not normal, maybe log or polynomial?
reg_sturgeon2 <- lm(log(avg_count) ~ species_total,
                   data = sturgeon_exp)
summary(reg_sturgeon2)
plot(reg_sturgeon2)

```


```{r test}

ggplot(sturgeon_exp, aes(year)) +
  geom_smooth(aes(y = avg_count,
                  color = 'red'),
              se = F) +
  geom_smooth(aes(y = species_total/10000000,
                  color = 'blue'),
              se = F) +
  scale_x_discrete(limits = seq(2002,2016,2)) +
  scale_color_discrete(name = "Key",
                       labels = c("Count", "Expenditure\n($10,000,000s)")) +
  labs(title = 'Sturgeon expenditure and population counts',
       x = 'Year',
       y = 'Expenditure and average counts')

```
