---
title: "Project stuff"
author: "Cole Hunter"
date: "3/19/2021"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Loading in 
```{r Load, include=FALSE}
library(tidyverse)
library(mapview)
library(leaflet)
library(sf)
library(sp)
# library(rnaturalearth)
# library(rnaturalearthdata)
setwd("C:/Users/wmjhu/Downloads")
animals <-read_csv("LPR2020data_public.csv")
```
##NA's
```{r adjust NA}
#Getting all the null values as NA
animals[apply(animals, 2, function(x) x=="NULL")] = NA

percent_na <- sum(is.na(animals))/(15348*98)

cat("Given",percent_na,"percent of the data is stored as NA values,
    it would be helpful to know if the unknown values are common to the rows or 
    columns")

# colSums(is.na(animals))

na_col <- as_tibble(map(animals, ~sum(is.na(.)))) %>% select(30:98)
na_row <- na_col %>% pivot_longer(cols =1:69 ,names_to = "Year", values_to = "Missing")


# Graph of missing data percentage for time series portion of data
na_row %>%
  filter(Missing!=0) %>%
  ggplot(aes(x = Year, y = Missing/nrow(animals), fill = Missing/nrow(animals) ))+
  geom_bar(stat = 'identity')+
  scale_fill_viridis_b()+
  labs(title = "Percent of Missing Time Series Data by Year",y = "", fill = "Percent Missing")

#Percent of time series data missing
missing_ts <- sum(na_row$Missing)/(nrow(animals)*69)
print(missing_ts)
```
A big problem with the data set is that it has a lot of missing values for the time series data. From the graph above, we can see that there is less missing data between the mid 1980's and early 1990's, but towards the beginning of when there stats were tracked, as well as in more recent times, the amount of data available for us to track changes over time is extremely limited. With this in mind, it seems like it would be best to try and see what trends we can observe in the data, and use that to target a new data set to analyze that contains more data about animal population trends, perhaps in certain regions. 


```{r Maps}
#Taking a look at the types of animals
animals %>%
  ggplot(aes(y = Class, fill = T_biome))+
  geom_bar()+
  labs(title = "Number of Animals by Class and Biome Type", xlab = "Count",
       ylab = "Class", fill = "Climate")
cat("From the plot we can see that the most represented animals in the data set are mammals, fish, birds, mammals and reptiles.")

#Map data
#Need to get the latitude values into the correct format, remove trailing 0's
animals$Latitude <- as.numeric(str_remove(animals$Latitude, "0+"))

#Splitting into the most represented Classes
mammals <- animals %>% filter(Class == "Mammalia") %>% select(ID,Latitude,Longitude)
reptiles <- animals %>% filter(Class == "Reptilia") %>% select(ID,Latitude,Longitude)
birds <- animals %>% filter(Class == "Aves") %>% select(ID,Latitude,Longitude)
fish <- animals %>% filter(Class == "Actinopteri") %>% select(ID,Latitude,Longitude)
fish <- na.omit(fish)
mammals <- na.omit(mammals)
birds <- na.omit(birds)
reptiles <- na.omit(reptiles)

# All of the collection points
animal_loc <- animals %>%
  filter(!is.na(Latitude),!is.na(Longitude), Specific_location == "Yes") %>%
  select(ID, Class, Latitude, Longitude)
mapview(animal_loc, xcol = "Longitude", ycol = "Latitude", crs = 4326,layer.name = "Location of Animal Counts")

#Plotting the collection location on a map
mammal_map <- mapview(mammals,xcol = "Longitude", ycol = "Latitude", crs = 4326, grid = FALSE,layer.name = "Location of Mammal Counts")
reptile_map <- mapview(reptiles,xcol = "Longitude", ycol = "Latitude", crs = 4326, grid = FALSE,layer.name = "Location of Reptile Counts")
bird_map <- mapview(birds,xcol = "Longitude", ycol = "Latitude", crs = 4326, grid = FALSE,layer.name = "Location of Bird Counts")
fish_map <- mapview(fish,xcol = "Longitude", ycol = "Latitude", crs = 4326, grid = FALSE,layer.name = "Location of Fish Counts")

major_maps <- mammal_map+reptile_map+bird_map+fish_map
major_maps

# Get an idea of where the bulk of the data points are collected
animals %>% ggplot(aes(Latitude, fill = Class))+geom_histogram()+
  labs(title = "Location of Data Collection")
# Can get an idea of where the 4 most common Classes are located
animals %>%
  filter(Class == c('Mammalia','Reptilia','Actinopteri','Aves')) %>% 
  ggplot(aes(Latitude, fill = Class))+geom_histogram()+
  labs(title = "Location of Data Collection for Top 4 Classes")

#Majority of the collected species are located in the Northern Hemisphere, check the actual value

nhem_per <- sum(animals$Latitude>=0,na.rm = TRUE)/length(!is.na(animals$Latitude))
print(nhem_per)

```
From the above graphs, we can see that the collection sites of animals is mostly concentrated in the northern hemisphere. In fact, out of the 15,348 animals that have been counted in this data set, 76 percent of those were located in the northern hemisphere. 

Its also interesting to note that the graphs for reptiles is the one that shows the tightest spread of count locations relative to the equator. 

It would be helpful to get a sense of how this data lines up relative to the general location of collections.

```{r stuff}
top_ten_loc  <- animals %>% group_by(Location) %>% summarize(count = n()) %>%  arrange(desc(count)) %>% head(10)
top_ten_loc

top_ten_country <-animals %>% group_by(Country) %>% summarize(count = n()) %>%  arrange(desc(count)) %>% head(10)
top_ten_country


# Looking for only the counts that are individual counts
only_ind <- animals %>% filter(Units == "Number of Individuals" |
                       Units == "Number of individuals" |
                       Units == "individuals" |
                       Units == "Individuals")
missing_ind <- (sum(is.na(only_ind))/(nrow(only_ind)*ncol(only_ind)))
missing_ind

```
Even after filtering for those counts which only represent some form of an individual count for the given animal, the data still contains missing values for over 67 percent of its values. 

Looking at the two tables above, it might help to know if the regions that have the most animals tracked are also the regions that have more of the time series data available, as this could help us focus in on the most well documented population trends. 

```{r only individuals counts}
animal_individual <- animals %>% filter(str_detect(Units, "individuals|Individuals|individual|Individual"))


(sum(is.na(animal_individual))/(nrow(animal_individual)*ncol(animal_individual)))
animal_individual <- animal_individual %>% pivot_longer(cols = '1950':ncol(animal_individual),names_to = "Year", values_to = "Count")
# Dropping rows that do not have a count for the given year
animal_individual<-animal_individual[complete.cases(animal_individual[31]),]
```
At this point we have all of the population data for each all of the animals.
To go further, we can group by the ID number and then arrange in descending order to find the animals that have the most number of years with population data available.
```{r Most years available}
animal_individual <- animal_individual %>% group_by(ID) %>% mutate(no_years = n())
most_year <- animal_individual %>%
  group_by(ID,Common_name,Country) %>%
  summarize(no_years = n()) %>%
  arrange(desc(no_years)) %>% 
  head(40)

# See how many of the top 40 tracked animals are located in each country
common_country<-most_year %>% group_by(Country) %>% summarize(count = n()) %>% arrange(desc(count))
```
From the table above we see that there are a number of species that have been tracked quite well throughout the United States and Canada. It's also  important to note that the same species can show up more than once, like Sockeye salmon, if they are being tracked at multiple locations. Given the data shows that the most well tracked animal populations are those that are located in the United States and Canada, we can focus in on those particular regions, and look at the tracked species to try and find trends in their populations over the years. 
```{r other}


# Need to see how many different methods of counting are used in the data set
# count_methods <- c(unique(animals_adj$Method))
# cat(length(count_methods))
# 
# 
# animals_adj <- animals %>%
#   pivot_longer(cols = '1950':ncol(animals), names_to = "Year", values_to = "Counts") %>%
#   filter(!is.na(Counts))

# animals_adj %>% group_by(ID) %>% filter(Counts<10000) %>% ggplot(aes(x = Year, y = Counts ))+
#   geom_point(stat = 'identity')
```



