<<<<<<< HEAD
---
title: "Initial LP Data Analysis"
output: html_notebook
author: "Katie"
---

Notebook for initial look at the Living Planet data, saved as `LPR2020.csv` in `data`.  The working directory should be `final-project-bald_eagle/code`.

```{r setup}
# check working directory
getwd()

# packages
suppressPackageStartupMessages(library(tidyverse))
```

Importing data from CSV.  ID is imported as integer; latitude, longitude, and all of the count data as double; everything else as character.  Note that some of the features should be factors, but that will be taken care of later, to keep control of factor levels.

```{r import}

LPI_data <- read_csv('../dataset/LPR2020.csv',
                     col_types = 
                       cols_only(ID = col_integer(),
                                 Binomial = col_character(),
                                 Class = col_character(),
                                 Order = col_character(),
                                 Family = col_character(),
                                 Genus = col_character(),
                                 Species = col_character(),
                                 Subspecies = col_character(),
                                 Common_name = col_character(),
                                 Location = col_character(),
                                 Country = col_character(),
                                 All_countries = col_character(),
                                 Region = col_character(),
                                 Latitude = col_double(),
                                 Longitude = col_double(),
                                 Specific_location = col_character(),
                                 temperate_or_tropical = col_character(),
                                 System = col_character(),
                                 T_realm = col_character(),
                                 T_biome = col_character(),
                                 FW_realm = col_character(),
                                 FW_biome = col_character(),
                                 M_realm = col_character(),
                                 M_ocean = col_character(),
                                 M_biome = col_character(),
                                 Units = col_character(),
                                 Method = col_character(),
                                 `1950` = col_double(),
                                 `1951` = col_double(),
                                 `1952` = col_double(),
                                 `1953` = col_double(),
                                 `1954` = col_double(),
                                 `1955` = col_double(),
                                 `1956` = col_double(),
                                 `1957` = col_double(),
                                 `1958` = col_double(),
                                 `1959` = col_double(),
                                 `1960` = col_double(),
                                 `1961` = col_double(),
                                 `1962` = col_double(),
                                 `1963` = col_double(),
                                 `1964` = col_double(),
                                 `1965` = col_double(),
                                 `1966` = col_double(),
                                 `1967` = col_double(),
                                 `1968` = col_double(),
                                 `1969` = col_double(),
                                 `1970` = col_double(),
                                 `1971` = col_double(),
                                 `1972` = col_double(),
                                 `1973` = col_double(),
                                 `1974` = col_double(),
                                 `1975` = col_double(),
                                 `1976` = col_double(),
                                 `1977` = col_double(),
                                 `1978` = col_double(),
                                 `1979` = col_double(),
                                 `1980` = col_double(),
                                 `1981` = col_double(),
                                 `1982` = col_double(),
                                 `1983` = col_double(),
                                 `1984` = col_double(),
                                 `1985` = col_double(),
                                 `1986` = col_double(),
                                 `1987` = col_double(),
                                 `1988` = col_double(),
                                 `1989` = col_double(),
                                 `1990` = col_double(),
                                 `1991` = col_double(),
                                 `1992` = col_double(),
                                 `1993` = col_double(),
                                 `1994` = col_double(),
                                 `1995` = col_double(),
                                 `1996` = col_double(),
                                 `1997` = col_double(),
                                 `1998` = col_double(),
                                 `1999` = col_double(),
                                 `2000` = col_double(),
                                 `2001` = col_double(),
                                 `2002` = col_double(),
                                 `2003` = col_double(),
                                 `2004` = col_double(),
                                 `2005` = col_double(),
                                 `2006` = col_double(),
                                 `2007` = col_double(),
                                 `2008` = col_double(),
                                 `2009` = col_double(),
                                 `2010` = col_double(),
                                 `2011` = col_double(),
                                 `2012` = col_double(),
                                 `2013` = col_double(),
                                 `2014` = col_double(),
                                 `2015` = col_double(),
                                 `2016` = col_double(),
                                 `2017` = col_double(),
                                 `2018` = col_double()),
                     na = 'NULL')
```

Looking for any rows with no count data.

```{r findnull}
LPI_data %>% 
  select(`1950`:`2018`) %>% 
  filter_all(all_vars(is.na(.)))

LPI_data %>% 
  select(`1950`:`2018`) %>% 
  filter_all(all_vars(!is.na(.)))
```

Note that no rows have no count data, and no rows have count data for all years.

Next, finding the proportions of existent count data for each row.

```{r proportions}
LPI_data %>%
  pivot_longer(cols = `1950`:`2018`,
               names_to = 'year',
               values_to = 'count') %>%
  group_by(ID) %>% 
  summarize(proportion = mean(!is.na(count))) %>% 
  ggplot(aes(x = proportion)) +
    geom_histogram(breaks = seq(0,1,1/30),
                   color = 'black',
                   fill = 'steelblue') +
    labs(title = 'amount of data per row',
         x = 'proportion of non-na')
```

A very large majority of the rows have count data for less than 25% of the years.  It may be worth focusing in on only rows with greater than a certain amount of count data.

Note that units are different for counts of each species.

```{r units-number}

LPI_data$Units %>% 
  unique() %>% 
  length()

LPI_data$Units %>% 
  str_to_lower() %>% 
  unique() %>% 
  length()

```

There are 1929 different units for counts; some of which are repeats.  Some of the units are different just on the basis of capitalization.

To try to get some standardization, a new dataframe, `cleaned` is made.  Note that the features have also been converted to lowercase for ease of typing.

```{r units-fix}

# copy for cleaning
cleaned <- LPI_data

# lowercase features for convenience
names(cleaned) <- tolower(names(cleaned))

# lowercase for units
cleaned$units <- str_to_lower(cleaned$units)

# list of unique units and count
unique_units <- cleaned %>%
  group_by(units) %>%
  count()

# filter to look for counts of individuals
unique_units %>%
  filter(str_detect(units, "ind"))

```

Many of the units are just number of individuals said in different ways.  This needs to be cleaned up.

















=======
---
title: "Initial LP Data Analysis"
output: html_notebook
author: "Katie"
---

Notebook for initial look at the Living Planet data, saved as `LPR2020.csv` in `data`.  The working directory should be `final-project-bald_eagle/code`.

```{r setup}
# check working directory
getwd()

# packages
suppressPackageStartupMessages(library(tidyverse))
```

Importing data from CSV.  ID is imported as integer; latitude, longitude, and all of the count data as double; everything else as character.  Note that some of the features should be factors, but that will be taken care of later, to keep control of factor levels.

```{r import}

LPI_data <- read_csv('../dataset/LPR2020.csv',
                     col_types = 
                       cols_only(ID = col_integer(),
                                 Binomial = col_character(),
                                 Class = col_character(),
                                 Order = col_character(),
                                 Family = col_character(),
                                 Genus = col_character(),
                                 Species = col_character(),
                                 Subspecies = col_character(),
                                 Common_name = col_character(),
                                 Location = col_character(),
                                 Country = col_character(),
                                 All_countries = col_character(),
                                 Region = col_character(),
                                 Latitude = col_double(),
                                 Longitude = col_double(),
                                 Specific_location = col_character(),
                                 temperate_or_tropical = col_character(),
                                 System = col_character(),
                                 T_realm = col_character(),
                                 T_biome = col_character(),
                                 FW_realm = col_character(),
                                 FW_biome = col_character(),
                                 M_realm = col_character(),
                                 M_ocean = col_character(),
                                 M_biome = col_character(),
                                 Units = col_character(),
                                 Method = col_character(),
                                 `1950` = col_double(),
                                 `1951` = col_double(),
                                 `1952` = col_double(),
                                 `1953` = col_double(),
                                 `1954` = col_double(),
                                 `1955` = col_double(),
                                 `1956` = col_double(),
                                 `1957` = col_double(),
                                 `1958` = col_double(),
                                 `1959` = col_double(),
                                 `1960` = col_double(),
                                 `1961` = col_double(),
                                 `1962` = col_double(),
                                 `1963` = col_double(),
                                 `1964` = col_double(),
                                 `1965` = col_double(),
                                 `1966` = col_double(),
                                 `1967` = col_double(),
                                 `1968` = col_double(),
                                 `1969` = col_double(),
                                 `1970` = col_double(),
                                 `1971` = col_double(),
                                 `1972` = col_double(),
                                 `1973` = col_double(),
                                 `1974` = col_double(),
                                 `1975` = col_double(),
                                 `1976` = col_double(),
                                 `1977` = col_double(),
                                 `1978` = col_double(),
                                 `1979` = col_double(),
                                 `1980` = col_double(),
                                 `1981` = col_double(),
                                 `1982` = col_double(),
                                 `1983` = col_double(),
                                 `1984` = col_double(),
                                 `1985` = col_double(),
                                 `1986` = col_double(),
                                 `1987` = col_double(),
                                 `1988` = col_double(),
                                 `1989` = col_double(),
                                 `1990` = col_double(),
                                 `1991` = col_double(),
                                 `1992` = col_double(),
                                 `1993` = col_double(),
                                 `1994` = col_double(),
                                 `1995` = col_double(),
                                 `1996` = col_double(),
                                 `1997` = col_double(),
                                 `1998` = col_double(),
                                 `1999` = col_double(),
                                 `2000` = col_double(),
                                 `2001` = col_double(),
                                 `2002` = col_double(),
                                 `2003` = col_double(),
                                 `2004` = col_double(),
                                 `2005` = col_double(),
                                 `2006` = col_double(),
                                 `2007` = col_double(),
                                 `2008` = col_double(),
                                 `2009` = col_double(),
                                 `2010` = col_double(),
                                 `2011` = col_double(),
                                 `2012` = col_double(),
                                 `2013` = col_double(),
                                 `2014` = col_double(),
                                 `2015` = col_double(),
                                 `2016` = col_double(),
                                 `2017` = col_double(),
                                 `2018` = col_double()),
                     na = 'NULL')
```

Looking for any rows with no count data.

```{r findnull}
LPI_data %>% 
  select(`1950`:`2018`) %>% 
  filter_all(all_vars(is.na(.)))

LPI_data %>% 
  select(`1950`:`2018`) %>% 
  filter_all(all_vars(!is.na(.)))
```

Note that no rows have no count data, and no rows have count data for all years.

Next, finding the proportions of existent count data for each row.

```{r proportions}
LPI_data %>%
  pivot_longer(cols = `1950`:`2018`,
               names_to = 'year',
               values_to = 'count') %>%
  group_by(ID) %>% 
  summarize(proportion = mean(!is.na(count))) %>% 
  ggplot(aes(x = proportion)) +
    geom_histogram(breaks = seq(0,1,1/30),
                   color = 'black',
                   fill = 'steelblue') +
    labs(title = 'Proportion of non-NA data per row',
         x = 'Proportion',
         y = 'Row count')
```

```{r proportions-filtered}

LPI_data %>%
  pivot_longer(cols = `1950`:`2018`,
               names_to = 'year',
               values_to = 'count') %>%
  filter(!is.na(count)) %>%
  ggplot(aes(as.numeric(year))) +
    geom_density() +
    labs(title = 'Density of non-NA data by year',
         x = 'Year',
         y = 'Density')


```


A very large majority of the rows have count data for less than 25% of the years.  It may be worth focusing in on only rows with greater than a certain amount of count data.

Note that units are different for counts of each species.

```{r units-number}

LPI_data$Units %>% 
  unique() %>% 
  length()

LPI_data$Units %>% 
  str_to_lower() %>% 
  unique() %>% 
  length()

```

There are 1929 different units for counts; some of which are repeats.  Some of the units are different just on the basis of capitalization.

To try to get some standardization, a new dataframe, `cleaned` is made.  Note that the features have also been converted to lowercase for ease of typing.

```{r units-fix}

# copy for cleaning
cleaned <- LPI_data

# lowercase features for convenience
names(cleaned) <- tolower(names(cleaned))

# lowercase for units
cleaned$units <- str_to_lower(cleaned$units)

# list of unique units and count
unique_units <- cleaned %>%
  group_by(units) %>%
  count()

# filter to look for counts of individuals
unique_units %>%
  filter(str_detect(units, "ind"))

```

Many of the units are just number of individuals said in different ways.  This needs to be cleaned up.  We can use the `forcats` package.

Since that will be time-intensive to clean, a bit more analysis can be done on the rows with larger amounts of data to reduce the number of rows to work with.

```{r normalize}

cleaned <- cleaned %>% 
  select(id, binomial, latitude:specific_location, `1950`:`2018`)
  
cleaned[, 6:74] <- t(scale(t(cleaned[, 6:74])))

time_series <- cleaned %>% 
  pivot_longer(cols = `1950`:`2018`,
               names_to = 'year',
               values_to = 'count')

many <- cleaned %>% 
  filter(!is.na(count)) %>% 
  group_by(binomial) %>% 
  count() %>%
  ungroup() %>% 
  filter(n > 10)

time_series$year <- as.numeric(time_series$year)

time_series <- time_series %>% 
  filter(binomial %in% many$binomial)

ggplot(time_series, aes(as.numeric(year), count)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = 'lm')

reg_all <- lm(count ~ year, data = time_series)
  
  
```














>>>>>>> 2ed71247f43ec8a31d009b22a55f19f87e0a15a5
