---
title: "model"
author: "katie"
output: html_notebook
---

(Dataset code taken from our_data for consistency.)

```{r setup, include=FALSE}

# packages
library(tidyverse)
library(readxl)

```

## Current data files

**LPI_data**: original dataset from Living Planet

**LPI_only_obs**: Living Planet data set in time series format (each observation is a Year with a Count)

**redlist_data**: original dataset from IUCN Redlist

**merged_data**: merged dataset of LPI_data and redlist_data (not time series)

**merged_timeseries**: merged_data in time series format

**expen_X**: expenditure from FWS for Year X

**expenditures_X**: cleaned expenditure data for Year X

**expenditures_timeseries**: time series expenditure data (expenditures_200X compressed into one set)

**LPI_red_expen_ts**: time series dataset of LPI, IUCN Redlist, and Expenditures


## Generate Sets

## LPI_data
```{r LPI_data}

LPI_data <- read_csv('../dataset/LPR2020data_public.csv',
                     na = 'NULL',
                     col_types = 
                       cols_only(ID = col_factor(),
                                 Binomial = col_character(),
                                 Class = col_character(),
                                 Order = col_character(),
                                 Family = col_character(),
                                 Genus = col_character(),
                                 Species = col_character(),
                                 Subspecies = col_character(),
                                 Common_name = col_character(),
                                 Location = col_character(),
                                 Country = col_character(),
                                 All_countries = col_character(),
                                 Region = col_character(),
                                 Latitude = col_double(),
                                 Longitude = col_double(),
                                 Specific_location = col_character(),
                                 temperate_or_tropical = col_character(),
                                 System = col_character(),
                                 T_realm = col_character(),
                                 T_biome = col_character(),
                                 FW_realm = col_character(),
                                 FW_biome = col_character(),
                                 M_realm = col_character(),
                                 M_ocean = col_character(),
                                 M_biome = col_character(),
                                 Units = col_character(),
                                 Method = col_character(),
                                 `1950` = col_double(),
                                 `1951` = col_double(),
                                 `1952` = col_double(),
                                 `1953` = col_double(),
                                 `1954` = col_double(),
                                 `1955` = col_double(),
                                 `1956` = col_double(),
                                 `1957` = col_double(),
                                 `1958` = col_double(),
                                 `1959` = col_double(),
                                 `1960` = col_double(),
                                 `1961` = col_double(),
                                 `1962` = col_double(),
                                 `1963` = col_double(),
                                 `1964` = col_double(),
                                 `1965` = col_double(),
                                 `1966` = col_double(),
                                 `1967` = col_double(),
                                 `1968` = col_double(),
                                 `1969` = col_double(),
                                 `1970` = col_double(),
                                 `1971` = col_double(),
                                 `1972` = col_double(),
                                 `1973` = col_double(),
                                 `1974` = col_double(),
                                 `1975` = col_double(),
                                 `1976` = col_double(),
                                 `1977` = col_double(),
                                 `1978` = col_double(),
                                 `1979` = col_double(),
                                 `1980` = col_double(),
                                 `1981` = col_double(),
                                 `1982` = col_double(),
                                 `1983` = col_double(),
                                 `1984` = col_double(),
                                 `1985` = col_double(),
                                 `1986` = col_double(),
                                 `1987` = col_double(),
                                 `1988` = col_double(),
                                 `1989` = col_double(),
                                 `1990` = col_double(),
                                 `1991` = col_double(),
                                 `1992` = col_double(),
                                 `1993` = col_double(),
                                 `1994` = col_double(),
                                 `1995` = col_double(),
                                 `1996` = col_double(),
                                 `1997` = col_double(),
                                 `1998` = col_double(),
                                 `1999` = col_double(),
                                 `2000` = col_double(),
                                 `2001` = col_double(),
                                 `2002` = col_double(),
                                 `2003` = col_double(),
                                 `2004` = col_double(),
                                 `2005` = col_double(),
                                 `2006` = col_double(),
                                 `2007` = col_double(),
                                 `2008` = col_double(),
                                 `2009` = col_double(),
                                 `2010` = col_double(),
                                 `2011` = col_double(),
                                 `2012` = col_double(),
                                 `2013` = col_double(),
                                 `2014` = col_double(),
                                 `2015` = col_double(),
                                 `2016` = col_double(),
                                 `2017` = col_double(),
                                 `2018` = col_double()))
```

## LPI_only_obs
```{r LPI_only_obs}

LPI_only_obs <- LPI_data %>%
  pivot_longer(cols = `1950`:`2018`,
               names_to = "Year",
               values_to = "Count") %>%
  group_by(ID) %>%
  filter(!Count=='NULL') %>%
  mutate(Year = sub('.', '', Year))

LPI_only_obs$Year <- as.integer(LPI_only_obs$Year)

```

## redlist_data
```{r redlist_data}

redlist_data <- read_csv('../dataset/redlistspecies.csv',
                         col_types = cols(redlistCategory = col_factor()))

```

## merged_data
```{r merged_data}

redlist_data$scientificName <- gsub(" ", "_", redlist_data$scientificName)

# join livingplanet with redlist
merged_data <- left_join(LPI_data,
                         redlist_data,
                         by = c('Binomial'='scientificName'))

```

```{r merged_timeseries}

# pivot
merged_timeseries <- merged_data %>%
  pivot_longer(cols = `1950`:`2018`,
               names_to="Year",
               values_to="Count") %>%
  group_by(ID) %>%
  filter(!Count=='NULL') %>%
  mutate(Year=sub('.', '', Year))

# set classes
merged_timeseries$Year <- as.integer(merged_timeseries$Year)

```

## brRead

### These functions help read nested brackets for expenditure data
```{r brRead}

# These functions help read nested brackets
lastBrRead <- function(string){
   #lastBrRead reads the last outermost bracket
   #Example: "welcome (home) (my(lovely) gator)" %>% lastbrRead() returns "my(lovely) gator"
   openBr=0
   startpos=0
   endpos=0
   string_list <- unlist(strsplit(string, split = ""))
   for(pos in 1:nchar(string)){
      if(string_list[pos]=="("){
         if (openBr==0){
            startpos=pos
         }
         openBr<-openBr+1
      } else if(string_list[pos]==")"){
         if (openBr==1){
            endpos=pos
         }
         openBr<-openBr-1
      }
   }
   string<-substr(string,startpos+1,endpos-1)
   return(string)
}

rmBrRead<-function(string){
   if (nchar(string)==0) {return(string)}
   #rnBrRead removes any inner brackets
   #Example: "my(lovely) gator" %>% rmBrRead() returns "my gator"
   string_list<-unlist(strsplit(string, split = ""))
   openBr=0
   cutstartpos=0
   cutendpos=0
   for(pos2 in 1:nchar(string)){
      if(string_list[pos2]=="("){
         if (openBr==0){
            cutstartpos=pos2
         }
         openBr<-openBr+1
      } else if(string_list[pos2]==")"){
         if (openBr==1){
            cutendpos=pos2
         }
         openBr<-openBr-1
      }
   }
   string<-paste(substr(string, 1,cutstartpos-1), substr(string, cutendpos+1,nchar(string)), sep="")
   return(string)
}

brRead<-function(string){
   #"welcome (home) (my(lovely) gator)" %>% brRead() returns "my gator" 
   return(rmBrRead(lastBrRead(string)))
}
```

# expenditures_200X
```{r expenditure2017}
# Load data, cut rows, rename vars, set Classes
expenditure<-read_excel('../dataset/expen/2017-Expenditures-Report.xlsx')
expenditure$`Group Name`[is.na(expenditure$`Group Name`)]<-0 #set NA's as 0

expenditures_2017<-expenditure%>%
   rename(Class=`Group Name`, Species=`Species (50 CFR Part 17)`)%>% #rename variables
   mutate(temp=cumsum(Class=='Fishes subtotal'))%>% #set rows before "Fishes subtotal" to 0
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>% 
   mutate(temp=cumsum(!Class=="Group Name" & !sapply(Class,nchar)<=3))%>% #set Classes
   group_by(temp)%>%
   mutate(Class=dplyr::first(Class))%>%
   ungroup()%>%
   select(-temp)

#rename Class levels
expenditures_2017$Class<-as.factor(expenditures_2017$Class)
expenditures_2017$Class<-recode(expenditures_2017$Class, "Amphibians"="Amphibia", "Fishes"="Actinopteri", "Birds"="Aves", "Reptiles"="Reptilia", "Mammals"="Mammalia")

#Set Species/Subspecies/Location/Year (because bless them for putting 3 variables in one column)
expenditures_2017<-expenditures_2017%>%
   mutate(Species_loc=str_split(Species," - "))%>% #split by "-" into c(species,location)
   mutate(Species_sci=sapply(Species_loc, "[", 1), #get Species/Location cols
          Location=sapply(Species_loc, "[", 2))%>%
   select(-Species_loc)%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%relocate(Species, .after = Scientific)%>%
   mutate(Year=2017)

#Clean up data
expenditures_2017<-expenditures_2017%>%
   filter(!Species=="17") #indicating rows with headers
```

```{r expenditure2000}

expen_2000<-read_excel('../dataset/expen/expenditure_2000.xlsx',3)
expenditures_2000<-expen_2000%>%
   rename(Common_name="SPECIES NAME (50 CFR PART 17)", Scientific="SCIENTIFIC NAME", Class="RANK", FWS_total="FWS TOTAL", Other_fed="OTHER FED", Fed_total="FED TOTAL", States_total="STATES TOTAL", Species_total="SPECIES TOTAL ($000)")%>% #rename variables
   filter(!is.na(Common_name))%>%
   mutate(temp=cumsum(Common_name=='FISHES Subtotals'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Class),!is.na(Scientific))

expenditures_2000<-expenditures_2000%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>%
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA))%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   mutate(Year=2000)%>%
   mutate(FWS_total=FWS_total*1000, Other_fed=Other_fed*1000, Fed_total=Fed_total*1000, States_total=States_total*1000, Species_total=Species_total*1000)%>%
   data.frame()
```

```{r expenditure2001}
expen_2001<-read_excel('../dataset/expen/expenditure_2001.xlsx',5)

expenditures_2001<-expen_2001%>%
   rename(Species_sci="Species (50 CFR Part 17)", Class="...1", FWS_total="FWS\r\nTotal ($)", Other_fed="Other Fed ($)", Fed_total="Fed Total ($)", States_total="States Total ($)", Species_total="Species Total ($)")%>% #rename variables
   filter(!is.na(Species_sci))%>%
   mutate(temp=cumsum(Species_sci=='Fishes Subtotal'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Class))

expenditures_2001<-expenditures_2001%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   relocate(Species, .after = Scientific)%>%
   mutate(Year=2001)%>%
   data.frame()
```

```{r expenditure2002}
expen_2002<-read_excel('../dataset/expen/expenditure_2002.xlsx',5)

expenditures_2002<-expen_2002%>%
   rename(Species_sci="Species (50 CFR Part 17)", FWS_total="FWS\r\nTotal ($)", Other_fed="Other Fed ($)", Fed_total="Fed Total ($)", States_total="States Total ($)", Species_total="Species Total ($)")%>% #rename variables
   filter(!is.na(Species_sci))%>%
   mutate(temp=cumsum(Species_sci=='Fishes Subtotal'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Class))

expenditures_2002<-expenditures_2002%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   relocate(Species, .after = Scientific)%>%
   mutate(Year=2002)%>%
   data.frame()
```

```{r expenditure2003}
expen_2003<-read_excel('../dataset/expen/expenditure_2003.xlsx',5)

expenditures_2003<-expen_2003%>%
   rename(Species_sci="Species (50 CFR Part 17)", Status="Stat- us", FWS_total="FWS\r\nTotal ($)", Other_fed="Other Fed ($)", Fed_total="Fed Total ($)", States_total="States Total ($)", Species_total="Species Total ($)")%>% #rename variables
   filter(!is.na(Species_sci))%>%
   mutate(temp=cumsum(Species_sci=='Fishes Subtotal'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Class))

expenditures_2003<-expenditures_2003%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   relocate(Species, .after = Scientific)%>%
   mutate(Year=2003)%>%
   data.frame()
```

```{r expenditure2004}
expen_2004<-read_excel('../dataset/expen/expenditure_2004.xlsx',5)

expenditures_2004<-expen_2004%>%
   rename(Species_sci="Species (50 CFR Part 17)", FWS_total="FWS\r\nTotal", Other_fed="Other Fed", Fed_total="Fed Total", States_total="States Total", Species_total="Species Total")%>% #rename variables
   filter(!is.na(Species_sci))%>%
   mutate(temp=cumsum(Species_sci=='Fishes Subtotal'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Class))

expenditures_2004<-expenditures_2004%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   relocate(Species, .after = Scientific)%>%
   mutate(Year=2004)%>%
   data.frame()
```

```{r expenditure2005}
expen_2005<-read_excel('../dataset/expen/expenditure_2005_06.xlsx',1)

expenditures_2005<-expen_2005%>%
   rename(Species_sci="Species (50 CFR Part 17)", Class="Rank", FWS_total="FWS\r\nTotal", Other_fed="Other Fed", Fed_total="Fed Total", States_total="States Total", Species_total="Species Total")%>% #rename variables
   filter(!is.na(Species_sci))%>%
   mutate(temp=cumsum(Species_sci=='Fishes Subtotal'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Class))

expenditures_2005<-expenditures_2005%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   relocate(Species, .after = Scientific)%>%
   mutate(Year=2005)%>%
   data.frame()
```

```{r expenditure2006}
expen_2006<-read_excel('../dataset/expen/expenditure_2005_06.xlsx',2)

expenditures_2006<-expen_2006%>%
   rename(Species_sci="Species (50 CFR Part 17)", FWS_total="FWS\r\nTotal", Other_fed="Other Fed", Fed_total="Fed Total", States_total="States Total", Species_total="Species Total")%>% #rename variables
   filter(!is.na(Species_sci))%>%
   mutate(temp=cumsum(Species_sci=='Fishes Subtotal'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Class))

expenditures_2006<-expenditures_2006%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   relocate(Species, .after = Scientific)%>%
   mutate(Year=2006)%>%
   data.frame()
```

```{r expenditure2007}
expen_2007<-read_excel('../dataset/expen/expenditure_2007.xlsx',1)

expenditures_2007<-expen_2007%>%
   rename(Species_sci="Species (50 CFR Part 17)", FWS_total="FWS\r\nTotal", Other_fed="Other Fed", Fed_total="Fed Total", States_total="States Total", Species_total="Species Total")%>% #rename variables
   filter(!is.na(Species_sci))%>%
   mutate(temp=cumsum(Species_sci=='Fishes Subtotal'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Class))

expenditures_2007<-expenditures_2007%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   relocate(Species, .after = Scientific)%>%
   mutate(Year=2007)%>%
   data.frame()
```

```{r expenditure2008}
expen_2008<-read_excel('../dataset/expen/expenditure_2008.xlsx',2)%>%
   select(-starts_with("..."))

expenditures_2008<-expen_2008%>%
   rename(Species_sci="Species\r\n(50 CFR Part 17)",Class="Group\r\nName", FWS_total="FWS Total", Other_fed="Other Fed", Fed_total="Fed Total", States_total="States\r\nTotal", Species_total="Species\r\nTotal")%>% #rename variables
   filter(!is.na(Class))%>%
   mutate(temp=cumsum(Class=='Fishes\r\nsubtotal'))%>%
   filter(temp==0)%>% #select rows before "Fishes subtotal"
   select(-temp)%>%
   filter(!is.na(Species_sci))

#Set Species/Subspecies/Location/Year (because bless them for putting 3 variables in one column)
expenditures_2008<-expenditures_2008%>%
   rowwise%>% #subset by row
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),brRead(Species_sci),Species_sci))%>% #if there is no parentheses at all
   mutate(Scientific=ifelse(grepl("\\(",Species_sci),rmBrRead(Scientific),Scientific))%>% #if there is still parentheses after brRead
   ungroup()%>%
   mutate(Species=ifelse( #create Species variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3, #if Species is format: (Family Species Subspecies)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",2), #then return 2nd to last element (Species)
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),tail,1)))%>% #else return last element (Species)
   mutate(Subspecies=ifelse( #create Subspecies variable
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),length)==3,
      sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),"[",3),
      NA)
          )%>%
   mutate(Genus=sapply(strsplit(Scientific, "\\s{2,}| |\\)|\\(|\r\n"),head,1))%>%
   relocate(Species, .after = Scientific)%>%
   mutate(Year=2008)%>%
   data.frame()

#Clean up data
expenditures_2008<-expenditures_2008%>%
   filter(!Species=="17") #indicating rows with headers
```

## expnditures_timeseries

```{r expenditures_timeseries}
#Vertical join by columns (time series format)
expenditures_timeseries<-rbind(
   expenditures_2000%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year),
   expenditures_2001%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year),
   expenditures_2002%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year),
   expenditures_2003%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year),
   expenditures_2004%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year),
   expenditures_2005%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year),
   expenditures_2006%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year),
   expenditures_2007%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year),
   expenditures_2008%>%
      select(Class, Genus, Species, Subspecies, FWS_total, Other_fed, Fed_total, States_total, Species_total, Year)
   )
```

## LPI_red_expen_ts
```{r LPI_red_expen_ts}
#Coerce uniformity along join keys by setting to lowercase
merged_timeseries <- merged_timeseries %>%
   mutate(Class = tolower(Class),
          Species = tolower(Species),
          Subspecies = tolower(Subspecies),
          Genus = tolower(Genus))

expenditures_timeseries <- expenditures_timeseries%>%
   mutate(Class = tolower(Class),
          Species = tolower(Species),
          Subspecies = tolower(Subspecies),
          Genus = tolower(Genus))

#Set classes
expenditures_timeseries$FWS_total <- as.numeric(expenditures_timeseries$FWS_total)
expenditures_timeseries$Other_fed <- as.numeric(expenditures_timeseries$Other_fed)
expenditures_timeseries$Fed_total <- as.numeric(expenditures_timeseries$Fed_total)
expenditures_timeseries$States_total <- as.numeric(expenditures_timeseries$States_total)
expenditures_timeseries$Species_total <- as.numeric(expenditures_timeseries$Species_total)
expenditures_timeseries$Year <- as.double(expenditures_timeseries$Year)

LPI_red_expen_ts <- merged_timeseries %>%
  left_join(.,
            expenditures_timeseries,
            by=c("Year","Genus", "Species", "Class"))
```


Another model (justaa general linear)
```{r model}

suppressPackageStartupMessages(library(car))

reg_expen0 <- lm(FWS_total ~ Species_total + Year,
                data = expenditures_timeseries)

reg_expen1 <- lm(Species_total ~ Year,
                 data = expenditures_timeseries)

for_log0 <- expenditures_timeseries %>% 
  filter(Species_total != 0)

for_log1 <- expenditures_timeseries %>% 
  mutate(fixed_total = ifelse(Species_total == 0, 1, Species_total))

regl_expen0 <- glm(FWS_total ~ log(Species_total) + Year,
                   na.action = na.omit,
                   data = for_log0)

regl_expen0 <- glm(FWS_total ~ log(fixed_total) + Year,
                   na.action = na.omit,
                   data = for_log1)

summary(regl_expen0)

# scatterplotMatrix(~ FWS_total + Species_total + Year,
#                   data = expenditures_timeseries)

```

A few models with lag
```{r model}

lag <- expenditures_timeseries %>% 
  mutate(lag_year = Year - 5)

reg_expen2 <- lm(FWS_total ~ Species_total + (lag_year),
                data = lag)

summary(reg_expen2)
summary(reg_expen0)

```
