---
title: "LPI Code"
author: "Nolan Choy"
date: "3/17/2021"
output: github_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(tidyverse)
setwd('..')
LPI_data<-read.csv('./dataset/LPR2020data_public.csv')
```

## Pivot to time series format:

```{r timeseries}
LPI_only_obs<-LPI_data%>%
  pivot_longer(starts_with("X"),names_to="Year", values_to="Count")%>%
  group_by(ID)%>%
  filter(!Count=='NULL')%>%
  mutate(Year=sub('.', '', Year))

head(LPI_only_obs%>%select(ID,Common_name, Location, Year, Count),10)
```

Set a couple variables as factors or numeric
```{r setvars}
LPI_only_obs$ID<-as.factor(LPI_only_obs$ID)
LPI_only_obs$Count<-as.numeric(LPI_only_obs$Count)
```

## Mammal Plot

```{r USmammalplots}
LPI_only_obs%>%
  filter(Units=="Number of individuals", Class=='Mammalia', Country=="United States", !ID==2218)%>%
  ggplot(aes(x=Year,y=Count,group=ID,col=ID))+
  geom_line(aes(col=Common_name))+
  labs(title="U.S. Mammals (Units: Number of Individuals) without Walrus")
```  

Walrus was an outlier visually.

## Bird Plot

```{r USbirdsplot}
LPI_only_obs%>%
   filter(Units=="Number of individuals", Class=='Aves', Region=='North America')%>%
   ggplot(aes(x=Year,y=Count,group=ID,col=ID))+
   geom_line(aes(col=Common_name))+
   labs(title="North American Birds (Units: Number of Individuals)")
```

## Gorilla Plot

```{r Gorillaplot}
LPI_only_obs%>%
   filter(Common_name=="Western gorilla", Units=="Number of individuals"| Units=="Estimated number of individuals")%>%
   ggplot(aes(x=Year,y=Count,group=ID,col=ID))+
   geom_line(aes(col=Location))+
   labs(title="Western Gorilla Count (Units: (Estimated) Number of Individuals)")

LPI_only_obs%>%
   filter(Genus=="Gorilla", Units=="Number of individuals")%>%
   ggplot(aes(x=Year,y=Count,group=ID,col=ID))+
   geom_line(aes(col=Location))+
   labs(title="Mountain/Eastern Lowland Gorilla Count (Units: Number of Individuals)")
```  

# Merging with Redlist Data

Merging data from IUCN (notably, this data is from simple_summary.csv whereas data from assessment contains rationales for conservation status)

```{r merge}
setwd('..')
redlist_data<-read.csv('./dataset/redlistspecies.csv')
redlist_data$scientificName <- gsub(" ", "_", redlist_data$scientificName)

#join livingplanet with redlist
merged_data<-left_join(LPI_data,redlist_data, by=c('Binomial'='scientificName'))

#pivot
merged_timeseries<-merged_data%>%
  pivot_longer(starts_with("X"),names_to="Year", values_to="Count")%>%
  group_by(ID)%>%
  filter(!Count=='NULL')%>%
  mutate(Year=sub('.', '', Year))

#set types
merged_timeseries$ID<-as.factor(merged_timeseries$ID)
merged_timeseries$Count<-as.numeric(merged_timeseries$Count)
merged_timeseries$redlistCategory <- factor(merged_timeseries$redlistCategory, levels = c(NA, 'Data Deficient', 'Lower Risk/leask concern', 'Least Concern', 'Lower Risk/conservation dependent', 'Lower Risk/near threatened', 'Near Threatened', 'Vulnerable', 'Endangered', 'Critically Endangered', 'Extinct in the Wild', 'Extinct'))
merged_data$redlistCategory <- factor(merged_data$redlistCategory, levels = c(NA, 'Data Deficient', 'Lower Risk/leask concern', 'Least Concern', 'Lower Risk/conservation dependent', 'Lower Risk/near threatened', 'Near Threatened', 'Vulnerable', 'Endangered', 'Critically Endangered', 'Extinct in the Wild', 'Extinct'))
```

### US Mammal Plot by Redlist Status

```{r mergedUSmammalplot}
merged_timeseries%>%
  filter(Units=="Number of individuals", Class=='Mammalia', Country=="United States", !ID==2218)%>%
  ggplot(aes(x=Year,y=Count,group=ID,col=ID))+
  geom_line(aes(col=redlistCategory))+
  labs(title="U.S. Mammals (Units: Number of Individuals) without Walrus by RedlistConcern")
```


### Aves Order Conservation Category Distribution

```{r mergedGlobalbirdplot}
merged_timeseries%>%
   filter(Units=="Number of individuals" | Units=="Population estimate", Class=='Aves')%>%
   ggplot(aes(x=Year,y=Count,group=ID,col=ID))+
   geom_line(aes(col=redlistCategory))+
   labs(title="Global Birds (Units: Number of Individuals, Population Estimate) by RedlistConcern")
```

```{r AvespctDist}
merged_timeseries%>%
   filter(Class=="Aves")%>%
   group_by(redlistCategory,Order)%>%
   summarize(Category_count=n())%>%
   ungroup()%>%group_by(Order)%>%
   mutate(pct_Cat=Category_count/sum(Category_count))%>%
   ggplot(aes(Order,pct_Cat,fill=redlistCategory))+geom_col()+labs(title="Aves Order Conservation Category Distribution", y="Perecentage of Species in Category")
```


### Extinct Species list

```{r extinct}
merged_data%>%filter(redlistCategory=="Extinct")%>%select(c(Common_name))
```

Odd that Bridled white-eye is "Extinct" despite all other sources claiming their population to be Endangered.


## Class Conservation Category Distribution

```{r SpeciesDist}
merged_data%>%
   filter(!is.na(redlistCategory))%>%
   group_by(redlistCategory,Class)%>%
   summarize(Category_count=n())%>%
   ungroup()%>%group_by(Class)%>%
   mutate(pct_Cat=Category_count/sum(Category_count))%>%
   ggplot(aes(Class,pct_Cat,fill=redlistCategory))+geom_col()+labs(title="Class Conservation Category Distribution", y="Perecentage of Species in Category")
```

This could be because certain Classes have more species.